Distribution-oblivious Programming with Dish

### 2019-11-21 pldi/draft.tex

Can we find a solution for the commands in coreutils?

@@

We parse a shell script using a posix compliant parser that returns
the ast.

\kk{Text point: The AST Is not a very good data structure to
  manipulate for distribution, so we translate it to a kind of a
  dataflow graph.}

@@

\kk{Text point: Problem: Not all parts of a shell script are
  distributable, so we cannot arbitrarily distribute}

Even though commands are split in categories, and several pipelines
are distributable, not all parts of a shell script are stateless and
several of them have effects that could affect the behaviour of a
program after distribution.

\kk{Solution: We have found that distributing subtrees of the shell
  AST is safe. Can we have a semi-formal argument about that?}

Mention all the different constructs that our front end handles, and
why it is safe to add them to the distribution pipeline.

\subsection{Intermediate Representation/Dataflow graph}

Mention its constructs, and its semantics, and why it is a bit
different that standard dataflow graphs (where usually incoming edges
are either consumed with some arbitrary interleaving, or pair by
pair).

\kk{Text point: The main shell abstraction is files, which are batches
  or streams of characters (or lines) where the order matters. So we
  want a representation that captures order of inputs and outputs (and
  ours does because its node in the graph has a sequence of inputs.}

Explain that the intermediate representation represents workers that
communicate through files, and that it captures parallelism both in
the operator level, as well as data parallelism.

@@

\kk{Maybe this should be merged with the one above, especially if it
  talks about how distribution happens on the intermediate
  representation.}

\subsection{Scaling up the distributed computation/Vertically expanding the graph}

The planner acts on the dataflow graph, by first expanding it ``as
much as possible'' to expose parallelization opportunities. It uses
knowledge about the inputs of each stateless and pure command.

\kk{Explain the algorithm and how it pushes parallelization through
  the graph, while preserving correctness.}

\kk{Give an example}

@@

Future work on this: calling the shell shtepper to only partially
evaluate strings, just expand arguments, and not do any significant
computation.

@@

\begin{itemize}
\item Fault Tolerance
\item Distributed Planning: The planner at the moment is naive, and
  there is a huge literature on operator placement of distributed
  dataflow graphs that can be used to improve it.
\item Support of shell constructs: At the moment we have been
  conservative in the shell subsets that we handle, so that we don't
  introduce any unsafe parallelism. The point of this work was not to
  be able to handle a complete distributable subset of the shell, but
  rather a significantly big part of it, to demonstrate that there can
  be performance benefit from it.
\item Support distributing different classes: At the moment we are
  only dealing with stateless and pure commands, that is the ones that
  are most popular but also can offer the biggest benefits from
  parallelizing them. However, it would be interesting to explore the
  distributability of the file system commands, by using a distributed
  file system. THis would also require considerations about
  replication, consistency, and failures.
\end{itemize}

>>>>

\kk{Can we find a solution for the commands in coreutils?}

\kk{Important TODO: Find at least one more command that can be
  executed like sort. Does wc work? If there is no pure command that
  can be easily parallelized other than sort, this story is not very
  nice.}

\section{Intermediate Representation/Dataflow graph}
\label{ir}

\kk{Inline listing?}

\kk{Try to be more formal in the section, by using the notation that
  is defined in the nodes paragraph.}

In order to effectively apply optimizations and distribute a shell
script, it first has to be translated to a manipulable
representation. As mentioned in \ref{}, the two main abstractions of
the shell, are files that contain data, and commands that communicate
through these files. We propose a slight variation of the standard
dataflow graph, that is commonly found in stream processing systems,
as the representation of a distributed shell computation. The main
benefit of the dataflow graph model is that it clearly exposes
operator level parallelism \kk{Is this a common term? Is there any
  definitive citation for it?}, as different nodes can be independent
processing workers. In our case each node represents a command, and
each edge represents a file.

\kk{Before that I have to mention that the basic data type $D$ is a
  line of characters. If we move the edges section before the nodes
  that will be fine.}

\paragraph{Nodes - Commands}

Nodes $f$ of the graph represent functions from one (possibly empty)
input stream to an output stream $f : D^* \rightarrow D^*$.  This
representation captures the majority of shell commands. The nodes $f$
are monotonic, meaning that they cannot retract their output. More
formally, if $f(x) = y$, then for any $x'$, there exists $y'$ such
that $f(x.x') = y.y'$, where $.$ is standard concatenation and will be
omitted from now on. An example of a node is the following command
\kk{Does it need ref? Also maybe this example is too trivial and can
  be removed.}:

\begin{lstlisting}[language=sh, float=h, numbers=none, escapeinside={($}{$)}]
 grep ``foo''
\end{lstlisting}

\noindent
This command takes a stream of lines from its standard input and
returns a stream of lines on its standard output. An important
observation though is that commands in the shell might read their
input stream from a sequence of different sources. An illustrative
example is the command \kk{Inline listing}cat which reads its input
from a combination of files and its standard input.

\begin{lstlisting}[language=sh, float=h, numbers=none, escapeinside={($}{$)}]
 cat x - y
\end{lstlisting}

\kk{Should I reference that this example was takes from the cat
  manpage?} The example above reads file \texttt{x} until it
encounters the end of file, then reads from \texttt{stdin}, and then
reads from \texttt{y}. In order to handle this particularity, nodes in
the dataflow graph can have many incoming edges, that are clearly
ordered in a sequence. Note that this comes in contrast with standard
dataflow graph representations where multiple incoming edges either
mean zipping the input to pairs, or arbitraliy interleaving the
different inputs.

\kk{A note about nodes and commands that can write and read from more
  files that the ones in their streams}

Note that nodes can read from more files than just the files of their
stream (or respectively write to more files that their output stream),
however these files are static (e.g. used for configuration or logging
\kk{Would an example here help}) and do not represent streams of data,
and therefore are not considered part of the dataflow graph. Because
of the assumption in section Front End \ref{}, that each file exists
only once in the dataflow graph, we can safely assume that reading and
writing to these static files does not interfere with the distributed
implementation, and doesn't alter the behaviour of the program.

\paragraph{Edges - Files}

\kk{Maybe edges has to be moved before nodes, so that we can define
  the line data type}

Edges in the dataflow graph represent files, the basic data
abstraction of the shell. They are used as communication channels
between nodes in the graph, and as the input or output of the whole
graph. Edges are represented as possibly unbounded streams of type
$D*$, where $D$ represents a line, that is a sequence of characters
followed by \verb|\n| \kk{or EOF?}. They could either refer to a
specific named file in the file system, or just be FIFO pipes that
used for interprocess communication \kk{Are there any other file
  types, such as URLs, ...?}.

\kk{Show an example pipe and its graph}.

In order to expose operator parallelism with the output of a command,
files that don't point to a resource can have an upper bound on the
lines that they can tranfer. The following example illustrates why
this is essential.

\kk{An example of a command that has two outputs and one goes to one
  command, while another output goes to a different one. Maybe only
  show the graph here and not the command itself.}

On the other hand, some forms of data parallelism can be exposed when
knowing the size of the input files. As mentioned in \ref{} some pure
commands (such as cat -n) only need line information to become
stateless, and knowing the size of a file could allow the system to
split it in different chunks that can be processed independently. To
account for that, edges that refer to an input resource contain the
number of lines of the file that they refer to. \kk{At the moment this
  is not supported by our real system. The whole paragraph above can
  go if we decide that it is not a good idea to talk about it now.}

Finally, the edges that don't refer to a resource and do not start
from a node in the graph represent the graph input, while the edges
that don't refer to a resource and do not point to a node in the graph
are its outputs.

\kk{(Maybe) Depending on which of the two paragraphs goes last
  (nodes/edges), I have to explain there that the last incoming edge
  to a node must be unbounded (if it doesn't point to a resource). I
  am actually not sure whether I should include this here.}

\kk{(Maybe) It might be beneficial to mention that many inputs do not
  represent the same thing as many outputs. These two notions are not
  symmetric.}

\paragraph{Merge and Split}

\kk{Mergers and splitters, and how they are implemented using shell
  constructs (cat and head). This subsection might be a better fit in
  some other section, maybe something related to the planner or the
  implementation, as it is a very elegant way to implement splitters
  and mergers while staying completely inside the shell programming
  model.}

It is common in the dataflow graph for a node to have many incoming
edges, or many outgoing edges (when all but the last outgoing edges
are bounded). However some nodes, might read their input from one file
(e.g. stdin). In this case, the distributed implementation needs to
concatenate input files, and split output files. Here is an example:

\kk{Show an example graph! of cat | tr | sort. The command in the
  middle must only take input from one file (maybe just stdin).}

\kk{This transition needs to be stronger} Luckily, the shell offers
primitives for file manipulations that can be used to implement a
merger and a splitter. More concretely a merger can be implemented by
a simple \verb|cat|. Given a set of input files \verb|in1, in2, ...|
and an output file \verb|out|, a merger can be implemented as:

\begin{lstlisting}[language=sh, float=h, numbers=none]
 cat $in1 $in2 ... > $out
\end{lstlisting}

\noindent
On the other hand, given an input file \verb|in| that is a FIFO pipe,
and two output files \verb|out1, out2| the first of which is bounded
to \verb|N| lines, a 2-splitter can be implemented as:

\begin{lstlisting}[language=sh, float=h, numbers=none]
 head -n $N $in > $out1 ; cat $in > $out2
\end{lstlisting}

\noindent
Using the 2-splitter as a building block, a pipe can be split to
arbitrarily many different pipes.

\kk{Do I also need to include a splitter for normal files?}

@@

\kk{It might make sense to merge this section with the one below with
  the name: From Shell to IR and back}

As mentioned in \Cref{ir}, in order to generate the distributed
implementation of a shell script, \texttt{Dish} first translates it to
a dataflow graph. However, there are several components in shell
programs that cannot be arbitrarily parallelized without affecting the
program behaviour.

\kk{Give examples of writing to environment variables, producing
  irreversible side effects, the ; and \&\& operators.}

In order to address this issue, we introduce the notion of
distributable AST subtrees, and we design an analysis that given the
shell script Abstract Syntax Tree identifies distributable AST
subtrees and translates each one of them to a dataflow graph. The AST
of the shell script is produced using a POSIX compliant parser
\kk{Reference libdash github}.

@@

\kk{The goal of this section is to convince the reader ``intuitively''
  that we choose safe parts of a shell program to distribute. There is
  no formal argument, but one should be able to intuitively believe
  us.}

Consider the following example:

\kk{Give an example of a shell script that cannot be distributed
  completely, Possibly involving ; and writing to a variable.}

\kk{Intuitively explain what could be distributed and what not.}

\kk{Our notion of distributable subtrees has two dimensions. First,
  the state of the shell should be an invariant in a distributable
  subtree. We cannot know this completely without evaluating, so we do
  a conservative analysis. The second dimension is that there exists
  at least one component that can be abstracted as a node of the
  dataflow graph. This can be anything that includes a command.}

%% Even though commands are split in categories, and several pipelines
%% are distributable, not all parts of a shell script are stateless and
%% several of them have effects that could affect the behaviour of a
%% program after distribution.

\kk{All of our assumptions and study are based on the POSIX
  standard. Reference?}

\kk{Not checked Assumption. A command doesn't write/read to arbitrary
  files that are not mentioned in its arguments. This is necessary to
  enforce that no two commands concurrently read/write from the same
  file.}

\kk{Not checked Assumption: No signals, traps, and sets at the
  moment. At the moment we don't have a way to handle signals and
  traps.}

\kk{Partially checked Assumption. Only one reader and one writer to
  each file, and no cycles. OLD MORE CONSERVATIVE ASSUMPTION: The same
  file exists once in the substree. This is a conservative assumption
  that ensures that no two commands write or read to the same file
  concurrently. We kind of check it but not completely. We check that
  the expanded evaluated files are not the same, but one could have
  part of a file in a variable etc.}

\kk{Checked assumption: How can we make sure that nodes in the IR are
  deterministic. Do we make some assumption that enforces this? I
  guess that commands that are nodes in the IR are all deterministic?
  That is stateless and pure commands are deterministic.}

\paragraph{analysis}

The analysis is done in a DFS fashion on the AST. Distributable
subtrees grow from the bottom up, until a construct that doesn't
``allow'' them to go through is encountered.

%% Mention all the different constructs that our front end handles, and
%% intuitively mention why it is ``safe'' to add them to the distribution
%% pipeline.

Subtrees don't cross the following:

\begin{itemize}
\item Sequence, And,  or
\item Function definitions
\item Command Substitutions
\item control flow: while, for, if, case: While these constructs can
  be parallelized and there has been research on it, it is orthogonal
  to our work, and can be incorporated as future work.
\end{itemize}

On the other hand: A distributable subtree starts whenever there is a
pipe, or asynchronous list. If the subcomponents of the pipe or
asynchronous list were already graphs, then we connect them to the
bigger graph \kk{explain how we do pipe and asynchronous list}, and if
they were not graphs, we make each subcomponent a node in the subtree,
event though they are not commands. These nodes do not belong to any
category, but still have stdin and stdout, and can be executed
concurrently (as long as they satisfy the checked assumptions above),
since they don't introduce new behaviours, they would still be
executed concurrently when ran in the standard shell, producing
possibly non-deterministic results with different interleavings.

\kk{Explain how we merge two graphs when seeing a pipe}

Whenever we see a subtree of the AST that is a command, we check the
files that it writes to. If any of its arguments is a command
substitution or contains anything that is evaluated\kk{make this more
  precise}, its category is marked as sequential\kk{name?}. Otherwise
we check its its category and its input output arguments (using the
methods described in \Cref{ext}). The command is then initialized as
the node of a singular graph, with the inputs and outputs of the graph
being the ones of the node.

\kk{Maybe mention the recursive case. I am not sure if that should be
  here, or in the distribution planner. In any case, it should be
  after we say that an IR is replaced with a call to the planner with
  the IR saved in a file. The example is a command in a pipeline whose
  first argument is a command substitution with a pipeline. Executing
  the outside dataflow graph, would lead to the command to be executed
  as part of this shell. However its first argument would have a call
  to the planner itself, leading to another distributed implementation
  running in parallel. The rest of the pipeline would wait for this to
  end before it continues. Assuming that the files that both of these
  graphs use are different, and that the above assumptions hold, there
  should be no interference between the two graphs and they should be
  able to run in parallel}

@@

\kk{I think that there is not enough content for the planner section
  and it should probably just be called graph transformations. Then we
  could have a small subsection at the end about how the planner
  applies transformations iteratively starting from the source of the
  graph and pushing them until the end.}

As mentioned in \Cref{ir}, the dataflow graph exposes operator level
parallelism. Different nodes in the graph can be executed
concurrently, communicating through the edges. Operator level
parallelism is already exposed by the shell, by using pipes \texttt{|}
and the background operator \texttt{\&}. Unfortunately, simple
operator level parallelism leaves a lot to be desired, since scaling a
computation requires manually introducing more operators.

In order to adequately utilize the available computational resources,
a distributed implementation should exploit data parallelism. Data
parallelism is based on the observation that some operations produce
the same result when performed independently on subsets of the input
data, merging the results afterwards. Data level parallelism is the
basic optimization that is used by popular distributed processing
systems such as MapReduce\kk{cite} and Apache Spark\kk{cite}.

\subsection{Graph transformations}

In our case, this is achieved by iteratively applying graph
transformations that preserve correctness, while exposing
parallelization opportunities. We illustrate that with a motivating
example; consider the following program and its dataflow graph.

\begin{lstlisting}[language=sh, float=h, numbers=none, escapeinside={($}{$)}]
 grep ``foo'' f1 f2 // Add its graph
\end{lstlisting}

\noindent
In this example, \texttt{grep} is applied to files f1 and f2, and its
result is sent to standard output. However, the result would be the
same if we concatenated the results of \texttt{grep ``foo'' f1} and
\texttt{grep ``foo'' f2}. This intuition can be generalized to all
stateless commands, as they are all essentially ``linear'' with
respect to concatenation and the following equation holds.

\[
\forall f, \mathrm{stateless}(f) \implies f(x.x') = f(x).f(x')
\]

\kk{Complete this story with pure, after we have written the section
  about command categories.}

\subsection{Planner Algorithm}

The \texttt{Dish} planner extends the transformations that were
described above with edge splits in order to allow for the above
transformations to be performed. More precisely, it splits input files
and graph edges into several chunks so that they can be processed in
parallel by the next levels.

\kk{Explain the greedy algorithm that our planner performs. It splits
  input files, and then pushes parallelization through the graph,
  starting from the source nodes.}

\kk{Maybe give an example}

@@

Future work on this: calling the shell shtepper to only partially
evaluate strings, just expand arguments, and not do any significant
computation. Even better, we could be calling the shtepper just after
having parsed the ast, and decide on maximal distributable subtrees as
late as possible in the process. This would be awesome.

\section{Evaluation}

\kk{We should make sure to measure the time that it takes for the
  whole process to run (parsing, translating, optimizing, planning)
  together with the scripts. This will probably be negligible, but we
  still have to mention it. }

\kk{Also, we have to have one-two handcrafted examples that have more
  than one pipeline, to show the execution time end to end, as this
  cannot be shown with the one-pipeline examples since they only have
  one graph, so no back and forth between shell and dish.}

\kk{Furthermore it would be great to have one distributed example that
  runs in more than one node (if we have time)}

\kk{Also we have to make sure to measure with and without the final
  cat of output file, so that we show that this is also negligible.}

\kk{We have to make sure that we repeat that the results are the same
  in all tests that we run. Corectness, blah blah ...}

\kk{We should have at least some pipelines that have a bad pure
  command in them.}

@@

\begin{itemize}
\item Fault Tolerance
\item Distributed Planning: The planner at the moment is naive, and
  there is a huge literature on operator placement of distributed
  dataflow graphs that can be used to improve it.
\item Support of shell constructs: At the moment we have been
  conservative in the shell subsets that we handle, so that we don't
  introduce any unsafe parallelism. The point of this work was not to
  be able to handle a complete distributable subset of the shell, but
  rather a significantly big part of it, to demonstrate that there can
  be performance benefit from it.
\item Support distributing different classes: At the moment we are
  only dealing with stateless and pure commands, that is the ones that
  are most popular but also can offer the biggest benefits from
  parallelizing them. However, it would be interesting to explore the
  distributability of the file system commands, by using a distributed
  file system. THis would also require considerations about
  replication, consistency, and failures.
\item No cycles (multiple commands writing and reading from the same file)
\item Formal proof that the analysis indeed returns distributable
  subtrees. Previously there was no formal semantics for shell, but
  recent work (cite greenberg) allows us to make a formal proof that
  the analysis is indeed sound.
\end{itemize}
